\cleardoublepage

\phantomsection
\addcontentsline{toc}{section}{Acknowledgments}
{\LARGE\bf{Acknowledgments}}
\vskip 15mm

\normalfont\itshape
The past four years have truly been an incredible journey filled with tons of learning, challenges, and personal growth.
I am profoundly grateful for all the people who have marked this period -- both those who were already a part of my life and those who became new acquaintances, collaborators, and cherished friends along the way.
\vskip 5mm
Foremost, I extend my deepest gratitude to my supervisor Dragan for allowing me to pursue my passion in every step of the way, for providing unwavering support, and for guiding me in something I hold dear and cherish.
Thank you for revealing the wonderful world of science to me and thank you for making me a part of it.
\vskip 5mm
To my colleagues---Maja, Vicko, Mario, and Zoran---and my dear collaborators, Maxim and Giulia, for their immense support and invaluable guidance.
I learned a lot from you all.
However, a special thanks goes to my dearest colleague Anna.
From being one of my most favorite lecturers to becoming a life role model and a cherished friend, you hold a special place in my heart.
\vskip 5mm
I would also like to thank to the members of my dissertation committee---Zoran, Kun, Zvonimir, Vicko, and Mario---not only for their time, patience, and keen interest in my research, but also for their intellectual contributions that shaped my development as a scientist.
\vskip 5mm
To all my friends outside of the academia.
You all played a significant role in shaping the person I am today.
Without you, this journey would lack its meaning.
A special thanks goes to my brother-from-another-mother, my best friend, and my unwavering support, Marijo -- thank you for wonderful memories somehow always accompanied by delicious meals in the late hours.
And to Ana---my platonic soulmate---for every ``just a 20-minute'' coffee at the FESB cafe, for every word of encouragement, for understanding me, and for always being there.
\vskip 5mm
Finally, to my family.

To Anja, you have always been my inexhaustible source of motivation, my true and continuous support, my one and only love.
Thank you for standing by my side through every bitter and every sweet moment, for every conversation, for all the love.
Thank you for being mine. 

To my grandma. You are my guardian angel. 

And to my mom.
For you, there are no words worthy to describe the depth of gratitude, love, and respect I feel.
\vskip 5mm
I dedicate this thesis to my three beautiful ladies.
